The evolution of the transcriptome of mammalian placentas

### 2016-11-15 manuscript/manuscript.Rnw
tweak distribution plot

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Number of Genes Meeting FPKM Cutoff in N species")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10,linetype="dotted")
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=300,
                         color=`Min. Exp. in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=300,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species",
                                  keywidth=unit(3,"mm"),
                                  keyheight=unit(3,"mm")
                                  ))
      + theme_bw()
      + theme(legend.position=c(1,1),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),
              plot.margin=unit(c(0,0,0,0),"mm")
              )
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis
  shows the number of 1:1 orthologous genes included in the core
  placenta transcriptome if %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[1]]}
  %
  (or %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]},
  % 
  etc.) have a minimum expression in FPKM equal to the horizontal axis.
  For example, if the cutoff is set to 10 (vertical black line), 
  % 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '14 of 14',max(num_genes)]}
  %
  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
  least 10, %
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '13 of 14',max(num_genes)]}
  %
  1:1 orthologs would be included if 13 of 14 species have an
   with an FPKM of at least 10, and 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '10 of 14',max(num_genes)]}
  genes would be included if 10 of 14 species
  must have an FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Min. Exp. in` == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

@@

\subsubsection{Annexin Complexes}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

>>>>

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Number of Genes Meeting FPKM Cutoff in N species")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10,linetype="dotted")
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=300,
                         color=`Min. Exp. in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=300,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species"),
                                  keywidth=unit(3,"mm"),
                                  keyheight=unit(3,"mm"),
                                  title.hjust=0.5,
                                  label.position="right"
                                  ))
      + theme_bw()
      + theme(legend.position=c(0.98,0.98),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),
              plot.margin=unit(c(0,0,0,0),"mm")
              )
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis
  shows the number of 1:1 orthologous genes included in the core
  placenta transcriptome if %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[1]]}
  %
  (or %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]},
  % 
  etc.) have a minimum expression in FPKM equal to the horizontal axis.
  For example, if the cutoff is set to 10 (vertical black line), 
  % 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '14 of 14',max(num_genes)]}
  %
  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
  least 10, %
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '13 of 14',max(num_genes)]}
  %
  1:1 orthologs would be included if 13 of 14 species have an
   with an FPKM of at least 10, and 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '10 of 14',max(num_genes)]}
  genes would be included if 10 of 14 species
  must have an FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & 
`Min. Exp. in` == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

@@

\subsubsection{Immune Regulation}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.
